---
title: "AllInOne"
author: "Cole Barbes"
date: "2023-10-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(extraDistr)
library(BSDA)
```

```{r}
Binom_plotter<-function(n,p,lb,ub){
x<-seq(0,n)
hx<-dbinom(x,n,p)


i<-x>= lb & x<= ub

hxi<-dbinom(x[i],n,p)


plot(x,hx,type='h', xlim=c(0,n),ylim=c(0,max(hx)))


if (lb>=0&lb<=n){
  abline(v=lb,lty=3,col='blue')
}

if (ub>=0&ub<=n){
  abline(v=ub,lty=3,col='blue')
}
par(new=T)
plot(x[i],hxi,type='h', col="red", xlim=c(0,n),ylim=c(0,max(hx)),xlab="",ylab="" )

par(new=F)


area <- pbinom(ub, n, p) - pbinom(lb, n, p) + dbinom(lb,n,p)
result <- paste("P(",lb,"\u2264 X \u2264",ub,") =",
   signif(area, digits=3), "\n",
   "n=", n, " p=", p)
mtext(result,3)

}
```

## Author: Cole Barbes

## Pulling In data from a File

```{r}
data0 <- read.csv("US_STATES.csv")
```

## Check Normality

```{r}
hist(data0$LifeExp)
```

```{r}

library(car)

qqPlot(data0$LifeExp)
```

## Compute the Mean, Variance, and Standard Deviation

```{r}
mean(data0$LifeExp)

var(data0$LifeExp)

sd(data0$LifeExp)

```

## Quartiles from data set

In r,

p is usually for area under the curve and to the left of the cutting point - P(x\<=q]

q is the cutting point that p of the data will be to the left of that point

d is for discrete P(x=3)

binom for binomial

pois for poisson

norm for normal

## Normal Random Variable

Given p, find q

Given q, find p

Different P[.....]

```{r}
pnorm(q=15, mean=20, sd=)


```

## The Discrete Uniform Distribution

```{r}
# define the x-axis
x_uniform<-seq(1,30, length=30)

# calculate the uniform probabilities
y_uniform<- ddunif(x_uniform,min=1,max=30)

# plot the distribution
plot(x_uniform,y_uniform, type='h')
```

```{r}
ddunif()
```

## Binomial Random Variable

Assume the following random variable follows a binomial distribution with n = 20, p=0.6

Example 1: find P(x\<=13]

```{r}
pbinom(q=13,size=20,prob=0.6)
Binom_plotter(20, 0.6, 0, 13)
```

Example 2: find P(x\<12)

```{r}
pbinom(q=12,size=20,prob=0.6)
Binom_plotter(20, 0.6, 0, 11)
```

Example 3: find P(x=13)

```{r}

dbinom(13, 20, 0.6)
Binom_plotter(20, 0.6, 13, 13)
```

Example 4: find P(4\<X\<=13)

```{r}
pbinom(13, 20, 0.6)-pbinom(4, 20, 0.6)
Binom_plotter(20, 0.6, 5, 13)
```

Example 5: find P(4\<=X\<13)

```{r}
pbinom(13, 20, 0.6)-pbinom(4, 20, 0.6)
Binom_plotter(20, 0.6, 4, 13)
```

Example 6: find p(4\<=X\<=13)

Example 7: find P(4\<X\<13)

Example 8: find P(X\>=13)

Example 9: find P(x\>13)

### Homework

If x is a binomial random variable, the following probabilities represent each case

dbinom(xvalue, size, p or q)

a\. n= 5, x=1, p=0.2

```{r}
dbinom(1,5, .2)
```

b\. n= 4, x=2, q=0.4

`{r} dbinom(2, size=4, 0.4)}`

c\. n= 3, x=0, p=0.7

`{r} dbinom(0, 3, 0.7)}`

d\. n=5, x=3, p=0.1

`{r} dbinom(3, 5, 0.1)}`

b\. n=4, x=2, q=0.6

`{r} dbinom(2, 4, 0.6)}`

b\. n= 3, x=1, p=0.9

`{r} dbinom(1, 3, 0.9)}`

if x is a binomial random variable, find the following probabilities

a\. P(x\<10) for n=25, p=0.7

`{r} pbinom(10, 25, 0.7)}`

b\. P(x\>=10) for n=15, p=0.9

`{r} 1-pbinom(10, 15, 0.9)}`

c\. P(x=2) for n = 20, p= 0.2

`{r} dbinom(2, 20, 0.2)}`

## Poisson Random Variable

For a Poisson distribution the commands would be:

rpois(n,lambda)-Which generates a random sample from the Poisson distribution

dpois(x,lambda)-generates $P(X=x)$ for the Poisson distribution

ppois(q,lambda)- calculates $P(X\leq x)$ for a Poisson distribution.

qpois(p,lambda)- This gives the x value such that $P(X \leq x)=p$ for a Poisson distribution.

average number of weekly sightings is 2.6

```{r}

dpois(5, 2.6)
#qpois(0.95, 2.6)
ppois(2, 2.6)

lamb<-seq(0, 52)
x<-seq(0,52)
for (i in 1:length(lamb)){plot(x,dpois(x,lamb[i]),type="h",main=substitute(paste("n=36 p= ",a),list(a=lamb[i])))}
```

### Exponential

a\. theta = 6.25

```{r}
pexp(5, 1/6.25)
qexp(p=0.275, rate=1/6.25)
```

### Quiz 2

```{r}
pbinom(p=0.2,size=3, q=0.8)
Binom_plotter(3, 0.2, 0, 0.8)
```

```{r}
pbinom(q=0.8,size=3,prob=0.2)
Binom_plotter(3, 0.2, 0, 3)

dbinom(0, 3,  0.2)+dbinom(1, 3, 0.2)+dbinom(2, 3, 0.2)+dbinom(3, 3, 0.2)

dbim
```

## Note 17: Confidence Interval.

```{r}
CI_data <- function(data, PopSDknown, PopSDvalue, ConfCoef, digit ){
  
  # data should be a column of numeric data
  # PopSDknown should be "Yes" or "No", whether the Population standard deviation is known or not.
  # PopSDvalue should be the value of the pop sd value. If pop sd is NOT known, just input 0.
  # ConfCoef is the confidence coefficient, should be a real number between 0 and 1.
  # digit is the number of digits you want to present in the CI
  
  x_bar <- mean(data)
  n <- length(data)
  
  confLv <- ConfCoef*100
  
  if (PopSDknown == "Yes"){
        crval <- qnorm((1+ConfCoef)/2)
        SD <- PopSDvalue
        output1 <- paste("The population standard deviation is", bquote(.(PopSDvalue)))
        
  }
  else if (PopSDknown == "No"){
        crval = qt((1+ConfCoef)/2,n-1)
        SD <- sd(data)
        output1 <- paste("The population standard deviation is unknown.")
  }
  
  ci_ub <- round(x_bar + crval*SD/sqrt(n),digit)
  ci_lb <- round(x_bar - crval*SD/sqrt(n),digit)
  

  
  
  cat(output1,"\nThe sample size is", bquote(.(n)), ". The sample mean is", bquote(.(x_bar)), ". The sample standard deviation is", bquote(.(sd(data))),"\nThe", bquote(.(confLv)), "% Confidence Interval for the population mean is (", bquote(.(ci_lb)), bquote(.(ci_ub)), ").\n" )
  
}
```

```{r}
CI_stat <- function(x_bar,n, PopSDknown, SD, ConfCoef, digit ){
  
  # x_bar is the sample mean
  # n is the sample size
  # PopSDknown should be "Yes" or "No", whether the Population standard deviation is known or not
  # SD is the pop sd or sample sd, depends on the input of PopSDknown
  # ConfCoef is the confidence coefficient, should be a real number between 0 and 1.
  # digit is the number of digits you want to present in the CI
  

  confLv <- ConfCoef*100
  
  if (PopSDknown == "Yes"){
        crval <- qnorm((1+ConfCoef)/2)
        output1 <- paste("The population standard deviation is", bquote(.(SD)))
        
  }
  else if (PopSDknown == "No"){
        crval = qt((1+ConfCoef)/2,n-1)
        output1 <- paste("The population standard deviation is unknown.")
  }
  
  ci_ub <- round(x_bar + crval*SD/sqrt(n),digit)
  ci_lb <- round(x_bar - crval*SD/sqrt(n),digit)

  
  cat(output1,"\nThe sample size is", bquote(.(n)), ". The sample mean is", bquote(.(x_bar)), ". The sample standard deviation is", bquote(.(SD)),"\nThe", bquote(.(confLv)), "% Confidence Interval for the population mean is (", bquote(.(ci_lb)), bquote(.(ci_ub)), ").\n" )
  
}
```

### Example 1

```{r}
CI_stat(x_bar=120.1, n=50, PopSDknown="Yes", SD=20, ConfCoef=0.95, digit=4)
```

### Example 2

if you are given the data

```{r}
dataCI <- read.csv("Note17Example.csv")

CI_data(data=dataCI$Data, PopSDknown="No", PopSDvalue=0, ConfCoef=0.95, digit=4)
```

### Practice Exam

```{r}
# question 1 a die is found unfair. The probability distribution of the X. the outcome when you toss this die once is given in the following table
x<-c(1, 2, 3, 4, 5, 6)
p<-c(0.3, 0.1, 0.15, 0.15, 0.2, 0.1)

mu <- sum(x*p)
mu

var <- sum(x^2*p) - mu^2
var

sd = sqrt(var)
sd


```

```{r}

sd(c(0.3, 0.2, 0.15, 0.15, 0.2, 0.1))
```

```{r}

sum(dbinom(x=c(0, 1, 2), size=4, p=0.2))
```

```{r}

pnorm(q=15, mean=11, sd=4)-pnorm(q=10, mean=11, sd=4)

1-pnorm(q=20, mean=11, sd=4)

qnorm(p=0.90, mean=11, sd=4)
```

```{r}

pbinom(q=2, size=4, 0.2)

sum(dbinom(c(0, 1, 2), size=4, 0.2))
```

```{r}
dataPE <- read.csv("Exam2Prac7.csv")
```

```{r}
hist(dataPE$Sample.Data)

```

The above graph is approximately normal

```{r}
library(car)

qqPlot(dataPE$Sample.Data)

sd(dataPE$Sample.Data)
```

```{r}
CI_data(data=dataPE$Sample.Data, PopSDknown="No", PopSDvalue=0, ConfCoef=0.92, digit=4)
```

```{r}
pexp(q=5, rate=1/6.25)

qexp(p=0.275, rate=1/6.25)

1-pnorm(q=82, mean=80, sd=1)

1-pnorm(q=0.30, mean=0.27, sd=0.0444)
```

```{r}
dataE <- read.csv("Exam2.csv")
```

```{r}
hist(dataE$SampleData)
```

```{r}
library(car)

qqPlot(dataE$SampleData)

sd(dataE$SampleData)
```

```{r}
CI_data(data=dataE$SampleData, PopSDknown="No", PopSDvalue=0, ConfCoef=0.90, digit=4)
```

```{r}

g <- (89-75)/8
g

pnorm(q=69, mean=75, sd = 8) - pnorm(q=60, mean=75, sd = 8)

1-pnorm(q=90, mean=75, sd=8)

qnorm(p=0.90, mean=75, sd=8)
```

```{r}
dbinom(x=c(0, 1, 2, 3, 4, 5), size=5, p=0.3)

sum(dbinom(x=c(0, 1, 2), size=5, p=0.3))
```

```{r}
1-pnorm(q=23, mean=20, sd=2)

pnorm(q=20, mean=20, sd=2)

pnorm(q=23, mean=20, sd=2)-pnorm(q=16, mean=20, sd=2)

0.67*0.33

sqrt((0.67*(1-0.67))/1000)

1-pnorm(q=0.75, mean=0.67, sd=0.01487)


```

```{r}
pbinom(p=0.3,size=5, q=0.7)
Binom_plotter(5, 0.3,0.16, 0.00)

sqrt(5*0.21)
```

```{r}

0.3*0.7

sqrt(0.3*0.7*5)


SE <- pnorm(q=0.05,mean=0, sd=1)*(20/sqrt(50))

((pnorm(q=0.05,mean=0, sd=1)*20) / SE)^2

sqrt(5*0.3*0.7)

```

## Hypothesis testing

### Practice problems for class

### Question 1.

```{r}
# question 1

library(BSDA)

zsum.test(mean.x=542, sigma.x = 116, n.x=100, mu=515, alternative="greater", conf.level=0.95)

```

### H0 : $\mu$= 515

### Ha : $\mu$\>515

```{r}
# question 2
x <- c(11.4, 13.9, 11.2, 14.5, 15.2, 8.1, 12.4, 8.6, 10.5, 17.1, 9.8, 15.9)
var<-9
sd <- 3
alpha <- 0.01

library(car)
library(nortest)

qqPlot(x)

ad.test(x)


z.test(x=x, alternative="less", mu=15, sigma.x=3, conf.level=0.01)

```

```{r}
library(BSDA)

z.test(x=x,mu=15, sigma.x=3, alternative="less", conf.level=.99)
```

c

```{r}
# question 3
library(BSDA)
tsum.test(mean.x=2.2, s.x=6.1, n.x = 76, mu=0, alternative="greater", conf.level=(1-0.05))
```

```{r}

#question 4
x<-c(2.6, 3.2, 2.1, 3.0, 3.1, 2.9, 3.7, 2.8)
qqPlot(x)
ad.test(x)
t.test(x=x,mu=3.5, alternative="two.sided", conf.level=0.95)
```

```{r}

x<-c(13, 108, 53, 45, 57, 44, 46, 58)

qqPlot(x)
ad.test(x)

# check symmetry with box plot
boxplot(x)

wilcox.test(x=x, mu=40, alternative="greater", conf.level=0.95)
```

```{r}
x<-c(75, 10, 92, 49, 58, 25, 96, 40, 90, 87, 10, 49, 20, 210, 34)

qqPlot(x)
ad.test(x)

boxplot(x)

SIGN.test(x=x, md=40, alternative="greater", conf.leve=0.95)
```

```{r}

x<-c(29300, 41500, 40400, 38500, 43500, 37800, 69500, 41200, 38400, 59200)
y<-c(28800, 41600, 39800, 38500, 42600, 38000, 69200, 40100, 38200, 58500)
z<-c(500, -100, 600, 0, 900, -200, 300, 1100, 200, 700)


```

```         
Overall notes for quiz 3:
pvalue < alpha gives evidence to reject
pvalue >= alpha does not give enough evidence to confirm or deny either
Type 1 error: occurs if the researcher reject
probability of a type 1 error is given by alpha
type 2 error: 
```

## Quiz 3:

```{r}
# question 1
library(BSDA)
dataQ1 <- read.csv("SAT_Scores_100.csv")

x_bar <- mean(dataQ1$SampleData)

x_bar
sd <- 116

s <- sd(dataQ1$SampleData)
s

zsum.test(mean.x=x_bar, sigma.x=sd, n.x=100, mu=525, alternative="greater", conf.level=0.95)
```

```{r}
#question 2
dataQ2 <- read.csv("SAT_Scores_20.csv")

library(car)
library(nortest)

qqPlot(dataQ2$SampleData)

ad.test(dataQ2$SampleData)

x_bar <- mean(dataQ2$SampleData)
x_bar

s <- sd(dataQ2$SampleData)
s


tsum.test(mean.x=x_bar, s.x=s, n.x=20, mu=530, alternative="greater", conf.level=0.97)
```

## Exam 3 - Prac

```{r}
library(BSDA)
library(nortest)
library(carData)
library(car)

```

Q1:

```{r}
x1 <- c(62, 75, 170, 47, 47, 81, 58, 2, 43, 52)
ad.test(x1)
#p value < 0.05 so we cannot say that this is normal

qqPlot(x1)
## approximately normal
boxplot(x1)

SIGN.test(x=x1, y = NULL, md = 70, alternative="less", conf.level=0.90)
#check if symmetric mean approximately equals median > 0.05 then mean != median
abs(mean(x1)-median(x1))/abs(mean(x1))
```

Q2

```{r}

x21 <- c(210, 231, 219, 229, 223, 240, 315)
x22 <- c(126, 138, 127, 266, 174, 180, 147, 156, 148)

qqPlot(x21)
qqPlot(x22)
#since the x21 is less than n = 8
shapiro.test(x21)

ad.test(x22)

hist(x21)
hist(x22)


wilcox.test(x21, x22, mu=0, alternative="greater")
```

Q3:

```{r}

# Q3
x31 <- c(57, 68, 77, 63, 72, 83)
x32 <- c(73, 69, 87, 69, 86, 84)
d3 <- x31 - x32

shapiro.test(d3)

# get p value
t.test(x=d3, mu=0, alternative="less", conf.level=0.95)

# CI interval
t.test(x=d3, mu=0, alternative="two.sided", conf.level=0.97)
```

Q4:

```{r}
dataph <- read.csv("Exam3Prac4.csv")
x4 <- dataph$SampleData

ad.test(x4)

t.test(x=x4, mu=8.5, alternative="two.sided", conf.level=0.93)
```

Q5:

```{r}

# no data given just summary stats which leads us to use zsumtest since the data shows normal 
zsum.test(mean.x=548, sigma.x=116, n.x=80, mu=525, alternative="greater")

# CI
zsum.test(mean.x=548, sigma.x=116, n.x=80, mu=525, alternative="two.sided", conf.level=0.96)
```

# Exam 3

Q1

```{r}

x1 <- c(564, 498, 259, 303, 307)

qqPlot(x1)
shapiro.test(x1)

t.test(x=x1, mu=300, alternative="greater", conf.level = 0.95)
t.test(x=x1, mu=300, alternative="greater", conf.level = 0.90)
```

```{r}

x21 <- c(104, 82, 102, 96, 129, 89, 114, 107, 89, 103)
x22 <- c(103, 103, 91, 113, 102, 103, 92, 90, 114, 113)

dx <- x21 - x22

qqPlot(dx)

ad.test(dx)
dx

# pvalue from ad test shows greater than alpha so the sample is approximately normal 

t.test(x=dx, mu=0, alternative ="two.sided", conf.level=0.95)


```

Q3

```{r}

x31<-c(67, 68, 78, 75, 84)
x32<-c(59, 63, 81, 74, 78)

diff<- x31 - x32

shapiro.test(diff)

t.test(x=diff, mu=0, alternative="greater", conf.level=0.95)

t.test(x=diff, mu=0, alternative="two.sided", conf.level=0.97)
```

Q4

```{r}
datan <- read.csv("Exam3Prob4.csv")
xn <- datan$SampleData

ad.test(xn)

t.test(x=xn, mu=8.5, alternative="two.sided", conf.level=0.95)

t.test(x=xn, mu=8.5, alternative="two.sided", conf.level=0.93)
```

Q5

```{r}

zsum.test(mean.x=4.98, sigma.x=1.62, n.x=258, mu=4.7, alternative="greater", conf.level=0.95)


zsum.test(mean.x=4.98, sigma.x=1.62, n.x=258, mu=4.7, alternative="two.sided", conf.level=0.96)
```

# Final Exam

Q

Q22

```{r}
testData <- read.csv("FinalPrac22.csv")

X <- testData$X
Y <- testData$Y

mod <- lm(Y~X)
summary(mod)

anova(mod)
confint(mod, level=0.95)

cor(X,Y)

predict(mod, newData=testData.frame(X=1.25), level=0.94, interval="confidence")
predict(mod, newData=testData.frame(X=1.25), level=0.94, interval="predict")

```
